### 2013-07-23 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
working on frontmatter
finished add step problem to nonlinear chapter

%%---------------------------------------------------------------------------%%
\section{Summary}
\label{sec:parallel_summary}

In this chapter we have developed a domain-decomposed parallel
algorithm for Monte Carlo synthetic acceleration and explored the
qualities of the algorithm through parallel scaling studies. The
following are the signficant observations and findings.

\begin{itemize}
\item MCSA can be effectively parallelized using domain decomposition
\item The leakage fraction of stochastic histories from a parallel
  subdomain can be quantified analytically using algebraic quantities
\item The multiple-set overlapping-domain (MSOD) parallel algorithm
  for particle transport has been adapted to parallelize MCSA
\item MCSA parallelized with MSOD was verified to produce the same
  results for solutions to the neutron diffusion equation as
  production Krylov methods
\item Strong scaling studies were performed with up to 65,536 cores
  and weak scaling studies were performed with up to 131,072 cores on
  the Titan Cray XK7 leadership class machine at Oak Ridge National
  Laboratory
\item MCSA scales favorably compared to production Krylov methods for
  both strong and weak scaling cases
\item Overlap in small quantities determined analytically by random
  walk length can provide parallel efficiency boosts of a few percent
  in strong scaling cases but is ineffective in weak scaling cases
\item Multiple sets will never boost parallel efficienicies in both
  weak and strong scaling cases due to additional communication
  overhead from added reduction operations
\item Multiple sets offers a means to reduce time to solution by
  solving multiple copies of the original problem and combining the
  solutions using superposition
\item MCSA is most efficiently used in parallel as a stochastic
  realization of an additive Schwarz method
\item With future optimization, large scale reactor analyses in which
  the transport operator is fully formed (e.g. the $SP_N$ equations)
  could experience improved time to solution on leadership-class
  machines with MCSA given the improved scaling performance
\end{itemize}

>>>>

%%---------------------------------------------------------------------------%%
\section{Summary\ }
\label{sec:parallel_summary}

In this chapter we have developed a domain-decomposed parallel
algorithm for Monte Carlo synthetic acceleration and explored the
qualities of the algorithm through parallel scaling studies. The
following are the significant observations and findings.

\begin{itemize}
\item MCSA can be effectively parallelized using domain decomposition
\item The leakage fraction of stochastic histories from a parallel
  subdomain can be quantified analytically using algebraic quantities
\item The multiple-set overlapping-domain (MSOD) parallel algorithm
  for particle transport has been adapted to parallelize MCSA
\item MCSA parallelized with MSOD was verified to produce the same
  results for solutions to the neutron diffusion equation as
  production Krylov methods
\item Strong scaling studies were performed with up to 65,536 cores
  and weak scaling studies were performed with up to 131,072 cores on
  the Titan Cray XK7 leadership class machine at Oak Ridge National
  Laboratory
\item MCSA scales favorably compared to production Krylov methods for
  both strong and weak scaling cases
\item Overlap in small quantities determined analytically by random
  walk length can provide parallel efficiency boosts of a few percent
  in strong scaling cases but is ineffective in weak scaling cases
\item Multiple sets will never boost parallel efficiencies in both
  weak and strong scaling cases due to additional communication
  overhead from added reduction operations
\item Multiple sets offers a means to reduce time to solution by
  solving multiple copies of the original problem and combining the
  solutions using superposition
\item MCSA is most efficiently used in parallel as a stochastic
  realization of an additive Schwarz method
\item With future optimization, large scale reactor analyses in which
  the transport operator is fully formed (e.g. the $SP_N$ equations)
  could experience improved time to solution on leadership-class
  machines with MCSA given the improved scaling performance
\end{itemize}

### 2013-07-23 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
finished add step problem to nonlinear chapter

Unfortunately, gaining convergence (and excellent iterative
performance) with MCSA for the $SP_N$ fuel assembly problem comes at
an immediate cost. Figure~\ref{fig:ilut_size} gives the maximum number
of non-zero entries in the composite linear operator generated by
preconditioning as a function ILUT drop tolerance for varying values
of fill level. For the 1-group $SP_1$ discretization, the original
linear operator will contain only a maximum of 7 non-zero entries per
row in the system. As observed in Figure~\ref{fig:ilut_size},
performing the explicit preconditioning yields composite linear
operators with $O(1,000)$ elements in a row for all combinations of
fill level and drop tolerance, over 10\% of the total row size for
this particular problem. This large number of row entries, observed
for a significant fraction of rows in the system, creates several
problems. First, sparsity is completely destroyed with each state in
the system now coupled to over 10\% of the total states in the system
through the composite iteration matrix. This means that Monte Carlo
sampling tables will be large, requiring significant memory to store
them and a substantial overhead in the sampling procedure during
transport. Second, the matrix-matrix multiply operations required to
build the composite operator see significant performance losses due to
the amount of data that must be handled. In parallel, losing sparsity
severely inhibits performance where now parallel operations require
communications amongst orders of magnitude more processors for
nearest-neighbor type algorithms.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/ilut_size.pdf}
  \end{center}
  \caption{\textbf{Maximum number of non-zero entries observed for all
      rows in the composite linear operator for the fuel assembly
      problem with ILUT preconditioning given as a function of ILUT
      drop tolerance.} \textit{Each colored curve represents the row
      size for a different ILUT fill level. Fill levels of 1.5, 2.0,
      2.5, and 3.0 were used.}}
  \label{fig:ilut_size}
\end{figure}

As a means to assess the quality of the preconditioning, a simple
metric is developed to address these concerns and allow comparison to
future developments. For our studies, our core performance metric will
be iterative performance with the minimum number of MCSA iterations
required to converge the problem desired. For Monte Carlo
calculations, this improved performance is balanced by the creation of
a dense composite system and we seek to reduce the number of non-zero
entries to a minimum value in order to lower the amount of coupling
amongst states in the system and reduce the amount of memory used
along with the potential latency overhead. Given these two objectives,
the following metric is proposed:
\begin{equation}
  \text{Quality Metric} = (\text{\# iterations}) \times (\text{maximum
    \# of non-zero values})\:,
\end{equation}
where the highest quality preconditioning is one that minimizes this
metric. For the ILUT preconditioned data provided,
Figure~\ref{fig:ilut_quality} provides the computed metric as a
function of ILUT drop tolerance for each fill level used. In general,
the metric is similar to the data observed for the number of
iterations required to converge as the non-zero row entries changed by
a smaller fraction over drop tolerances tested.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/ilut_quality.pdf}
  \end{center}
  \caption{\textbf{ILUT preconditioning quality metric for the fuel
      assembly problem given as a function of ILUT drop tolerance.}
    \textit{Each colored curve represents the quality metric behavior
      for a different ILUT fill level. Fill levels of 1.5, 2.0, 2.5,
      and 3.0 were used.}}
  \label{fig:ilut_quality}
\end{figure}

@@

In this chapter, the Monte Carlo synthetic acceleration method has
been applied to the $SP_N$ discretization of the neutron transport
equation and explored in the context of a difficult nuclear fuel
assembly criticality problem. The following are the signficant
observations and findings.

>>>>

Unfortunately, gaining convergence (and excellent iterative
performance) with MCSA for the $SP_N$ fuel assembly problem comes at
an immediate cost. Figure~\ref{fig:ilut_size} gives the maximum number
of non-zero entries in the composite linear operator generated by
preconditioning as a function ILUT drop tolerance for varying values
of fill level. For the 1-group $SP_1$ discretization, the original
linear operator will contain only a maximum of 7 non-zero entries per
row in the system. As observed in Figure~\ref{fig:ilut_size},
performing the explicit preconditioning yields composite linear
operators with $O(1,000)$ elements in a row for all combinations of
fill level and drop tolerance, over 10\% of the total row size for
this particular problem. This large number of row entries, observed
for a significant fraction of rows in the system, creates several
problems. First, sparsity is completely destroyed with each state in
the system now coupled to over 10\% of the total states in the system
through the composite iteration matrix. This means that Monte Carlo
sampling tables will be large, requiring significant memory to store
them and a substantial overhead in the sampling procedure during
transport. Second, the matrix-matrix multiply operations required to
build the composite operator see significant performance losses due to
the amount of data that must be handled. In parallel, losing sparsity
severely inhibits performance where now parallel operations require
communications among orders of magnitude more processors for
nearest-neighbor type algorithms.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/ilut_size.pdf}
  \end{center}
  \caption{\textbf{Maximum number of non-zero entries observed for all
      rows in the composite linear operator for the fuel assembly
      problem with ILUT preconditioning given as a function of ILUT
      drop tolerance.} \textit{Each colored curve represents the row
      size for a different ILUT fill level. Fill levels of 1.5, 2.0,
      2.5, and 3.0 were used.}}
  \label{fig:ilut_size}
\end{figure}

As a means to assess the quality of the preconditioning, a simple
metric is developed to address these concerns and allow comparison to
future developments. For our studies, our core performance metric will
be iterative performance with the minimum number of MCSA iterations
required to converge the problem desired. For Monte Carlo
calculations, this improved performance is balanced by the creation of
a dense composite system and we seek to reduce the number of non-zero
entries to a minimum value in order to lower the amount of coupling
among states in the system and reduce the amount of memory used
along with the potential latency overhead. Given these two objectives,
the following metric is proposed:
\begin{equation}
  \text{Quality Metric} = (\text{\# iterations}) \times (\text{maximum
    \# of non-zero values})\:,
\end{equation}
where the highest quality preconditioning is one that minimizes this
metric. For the ILUT preconditioned data provided,
Figure~\ref{fig:ilut_quality} provides the computed metric as a
function of ILUT drop tolerance for each fill level used. In general,
the metric is similar to the data observed for the number of
iterations required to converge as the non-zero row entries changed by
a smaller fraction over drop tolerances tested.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/ilut_quality.pdf}
  \end{center}
  \caption{\textbf{ILUT preconditioning quality metric for the fuel
      assembly problem given as a function of ILUT drop tolerance.}
    \textit{Each colored curve represents the quality metric behavior
      for a different ILUT fill level. Fill levels of 1.5, 2.0, 2.5,
      and 3.0 were used.}}
  \label{fig:ilut_quality}
\end{figure}

@@

In this chapter, the Monte Carlo synthetic acceleration method has
been applied to the $SP_N$ discretization of the neutron transport
equation and explored in the context of a difficult nuclear fuel
assembly criticality problem. The following are the significant
observations and findings.
